# *Monday*: Introduction to Survival Analyses and simulation of data {#survival}

## Key (operative) concepts {#key1}

1. Time has asymmetric density and can be censored:
  - not possible to summarize it by the means
  - cannot be normal distributed
  - use exponential family

2. Plot the log-plot to check the distribution assumptions

3. Censoring can be:
  - Right: event not (yet) occurred at f-up
    - Fixed (identical f-up for anyone)
    - Sequential ($min(T_i, C_i)$)
    - Random
  - Left: the event has occurred before the observed period (all population but
    not all information, e.g. menarche date)
  - Interval: the event can be occurred between two times (but don't know when)
  - Left truncated: starting point is after the beginning (different from Left,
    all the information but not complete population)

4. Models:
  - statistical: non-informative censoring (Kaplan-Meier, Cox model, ...)
  - probabilistic: independent censoring (life tables)
  - parametric (`survival::survreg()`, need to define the distribution) VS
    non-parametric (`survival::survfit()` or `rms::npsurv()`, no need to define
    distribution)


## Simulated Data {#sumulation1}

1. Simulate a sample of $n = 100$ or $1000$ exponential survival times, with
  mean $\theta = 5$.

  - Non censored

```{r simulation_par}
set.seed(171002)
n              <- c(thousand = 1000)                                   # samples
t              <- rexp(n, rate = 5)                   # random exponential times
status_no_cens <- rep(1, times = n)         # no censored data --> all are cases
```

  - Uniform censoring over $[0, a]$, with $a = 1, a = 0.5$ or $a = 2$

```{r}
a              <- c(cens_05 = 0.5)   # upper bound of the uniform censoring dist
cens           <- runif(n, min = 0, max = a)                    # censored times
t_cens         <- pmin(t, cens)    # censored times are earlier than event times
status_cens    <- status_no_cens - (t_cens == cens)      # remove censored cases
```




2. Plot the observed survival times

  - Non censored and censored
```{r sim_plots, fig.show='hold', out.width='50%'}
# NOTE: for the plots to be comparable, xlim and ylim have to be the same range
#       for both the plots. Moreover to drow well adjusted plots, they were set
#       a posteriori.
hist(t,
  main = 'Hystogram of uncensored times',
  col  = 'green',
  xlim = c(0, 1.5),
  ylim = c(0, 400),
  labels = TRUE                        # add the labels over the top of the bars
)
hist(t_cens,
  main = 'Hystogram of censored times (a = 0.5)',
  col  = 'red',
  xlim = c(0, 1.5),
  ylim = c(0, 400),
  labels = TRUE
)
```




3. Parametric estimation of survival function

  - Uncensored
  
```{r sim_survreg_nocens}
# `?survreg` := "Regression for a Parametric Survival Model"
# 
# R formula: y ~ x <--> math formula: y = f(x)
# 
# Here we want to model the response (labelled time) as they are, without any
# furter investigation on the effect on them from some other variable
survreg(Surv(t, status_no_cens) ~ 1,
  dist = 'exponential'
) %>%
  summary     # here `summary()` add some more statistics to the standard output
```

  - Censored
  
```{r sim_survreg_cens}
survreg(Surv(t_cens, status_cens) ~ 1,
  dist = 'exponential'
) %>%
  summary
```




4. Non parametric estimation of survival and the distribution functions

  - Uncensored

```{r survfit_nocens, fig.show='hold'}
# `?survfit` := "Create survival curves"
survfit(Surv(t, status_no_cens) ~ 1)

# Here we would like to compare to approach to survival plots:
# 1. Using the packege _survival_, so the standard one
# 2. Uisng the package _rms_, a comprehensive package for regression analyses

# Using survival `plot` provided by the _survival_ package
# (`?survival:::plot.survfit`), we can continue to
# use the `survfit()` function for nonparametric survival estimation from the
# same _survival_ package
survfit(Surv(t, status_no_cens) ~ 1) %>% 
  plot(
    xlim     = c(0, 1.55),
    conf.int  = TRUE,
    mark.time = TRUE,
    col       = 'green',
    main      = 'Uncensored --- survival'
)


# Using the survplot from the _rms_ package (`survplot`), we have to switch to
# the `npsurv()` function for nonparametric survival estimation from the _rms_
# package
npsurv(Surv(t, status_no_cens) ~ 1) %>% 
  survplot(
    xlim     = c(0, 1.5),
    conf.int = TRUE,
    n.risk   = TRUE,
    col      = 'green'
)
title(main = 'Uncensored --- rms') # unfortunally survplot do not have an
                                   # integrated option for the title...
```

  - censored
    
```{r survfit_cens, fig.show='hold'}
survfit(Surv(t_cens, status_cens) ~ 1)

survfit(Surv(t_cens, status_cens) ~ 1) %>%
  plot(
    xlim     = c(0, 0.55),
    conf.int  = TRUE,
    mark.time = TRUE,
    col      = 'red',
    main      = 'Censored (a = 0.5)'
)

npsurv(Surv(t_cens, status_cens) ~ 1) %>% 
  survplot(
    xlim     = c(0, 0.5),
    conf.int = TRUE,
    n.risk   = TRUE,
    col      = 'red'
)
title(main = 'Censored (a = 0.5) --- rms')
```








## `mgus` data from __survival__ package {#mgus}

1. Load and explore data

```{r mgus_explore}
data(mgus)                                                                # load
head(mgus)                                                       # first 10 rows
dim(mgus)                                              # number of rows and cols
names(mgus)                                                # name of the columns
str(mgus)                                   # R internal structure of the object

summary(mgus)                                              # summary from base R
 
describe(mgus)   # more comprehensive description from _Hisc_ package, loaded by
                 # the _rms_ one

mgus_df <- as_tibble(mgus)         # tidy data frame (important info printed all
                                   # together, and visualization auto-adjusted
                                   # to the consol width)
mgus_df
```




2. Non parametric Kaplan-Meyer estimation of the survival function

  - Estimate the survival function from randomization overall and
      according to sex.
    
```{r rmgus_survfit, fig.show='hold'}
survfit(Surv(futime, death) ~ 1,
  data = mgus_df
) %>%
  plot(
    conf.int  = TRUE,
    mark.time = TRUE,
    col       = 'blue',
    main      = 'Survival function for mgus data'
)


survfit(Surv(futime, death) ~ sex,
  data = mgus_df
) %>%
  plot(
    conf.int  = TRUE,
    mark.time = TRUE,
    main      = 'Survival function for mgus data according to sex',
    col       = c('red', 'blue'),
    lty       = c(2, 3)
)
legend(
  x = 10000, y = 1,
  legend = c("Female", "Male"),
  col    = c('red', 'blue'),
  lty    = c(2, 3)
)
```

```{r rmgus_survminer, fig.height=10}
# For survival object the package _survminer_ provide ggplot2 plots
# (`?ggsurvplot`) which could be very interesting and quite comprehensive.

survfit(Surv(futime, death) ~ sex,
  data = mgus_df
) %>% 
  ggsurvplot(
    conf.int            = TRUE,                      # draw confidence intervals
    pval                = TRUE,                                    # show pvalue
    pval.method         = TRUE,                            # print the test name
    title               = 'Survival curves for overall death according to sex.',
    xlab                = 'Days',
    legend              = 'right',                             # legend position
    legend.title        = 'Sex',
    legend.labs         = c('Female', 'Male'),
    risk.table          = TRUE,     # admits interesting options other than TRUE
    cumcensor           = TRUE,
    cumevents           = TRUE,
    pval.size           = 3.5,   # from here these are options passed to `ggpar`
    risk.table.fontsize = 3,     # for a better visualization
    fontsize            = 3,     # (auto-explicatives)
    xscale              = 30.44
  )
```

> Note: No female reaches the end of the f-up!

  - Test the effect of sex

```{r mgus_survdiff_and_rms}
# Using __survival__ (no plot method is provided for this solution)
survdiff(Surv(futime, death) ~ sex,
  data = mgus_df
)


# using __rms__
dd <- datadist(mgus_df)  # To evaluate cph, _rms_ needs this object which simply
                         # store statistics about the data.
                         # 
                         # Note: the name of the object (i.e. "dd") has to be 
                         #       exactly the same as the one specified into the
                         #       option set just after the `library(rms)` call.
                         #       (See: Chapter settings) 
cox_model <- cph(Surv(futime, death) ~ sex,
  data  = mgus_df
)

summary(cox_model)                           # return effect size and HR with CI

Predict(cox_model) %>%          # Compute predicted values and confidence limits
                                #
                                # Note: pay attention to Title-case "P"redict
  plot(
    groups = 'sex',
    anova  = anova(cox_model),       # Compute and print the $\chi^2$ statistics
    pval   = TRUE                    # print the pvalue 
)
```








## Non parametric Kaplan-Meier estimation of the survival function {#km1}

1. Let consider a sample of $n = 500$

```{r ex_ss}
n <- 500
```




2. Simulate the dates of entry in the cohort, from January, 2010 to January,
   2017

```{r ex_entry}
n_days <- 365.25 * 7              # Seven years, taking into account bissextiles
time_start <- runif(n = n,
  min = 0,
  max = n_days
) %>% 
  as.Date(origin = '2010-01-01')
```




3. Simulate the data-set of death, assuming exponential death times of mean $2$
   years

```{r ex_death}
mean_death_time <- 365.25 * 2
death_t         <- rexp(n, rate = 1 / mean_death_time)
status_no_cens  <- rep(1, n)
```




4. Let fix the reference date of the analyses of June, 2017

```{r ex_end}
end_date     <- as.Date('2017-06-01')           # Fixed date for the end of f-up
death_r_cens <- pmin(death_t, end_date - time_start)
status_cens  <- status_no_cens - (death_t == death_r_cens)
```




5. Estimate the survival function from randomization

```{r ex_surv_curves}
survfit(Surv(death_r_cens, status_cens) ~ 1) %>% 
  plot(
    conf.int  = TRUE,
    mark.time = TRUE,
    main      = 'Survival curve from randomization (right censored at 2017-06-01)',
    col       = 'blue',
    xlab      = 'Years',
    xscale    = 365.25
)
```

