# *Tuesday*: Cox models {#cox}


## Key (operative) concepts {#key2}

1. Non-informative censoring assumption!

> We cannot test for it, but we can be convinced of it

2. Test any covariates for proportional hazard. If fail:
  - If $H_0$ is valid, it is not a problem
  - Is it due to outliers?
  - Does this variable really need?
  - ... do you really think that proportional of hazard should hold? What about
    shift to a different model?
3. Test continuous variable for log-linearity. If fail:
  - try a transformation of the variable (i.e., log, spline, ...)
  - if it is not possible (e.g. *U-shape*) perform a categorization

> When performing categorization do not base it on the p-value: you have to 
  explain why this choiche is clinically relevant and not statistically 
  significant!

4. The biggest problem with databases with more observations for each patients
   is not the model but to produce a table with the right information in the
   right position. In particular we need the  following columns
   - id
   - start
   - end
   - event
   - covariates...

5. Get results easy to explain to / understand by a clinician!

## Basic tests and funtions {#test2}

For this part we will use the data `pbc` (`?pbc`) from the package __survival__.

> Note: `data(pbc)` load the `pbc` data-set and the `pbcseq` one, so on one side
>        we do not need to call `data(pbcseq)` to load the letter, on the other
>        side `data(pbcseq)` will throw an error because to load it we have to 
>        call `data(pbc)`. (We will use both data-sets.)

```{r data_load}
set.seed(171003)
data(pbc)                                                    # load the data-set
# ?pbc
pbc_df <- as_tibble(pbc)                       # create the tibble version of it
dd <- datadist(pbc_df)     # store in the dd variable its `datadist()` for _rms_

pbc_df                                                       # give a look at it
describe(pbc_df)                                 # and whatch at some statistics
```

### Impact of sex on death {sex2}

First of all we have to ask to our self, and to the clinicians, some questions:

1. There are non-informative censoring? Yes, because there is a final
   data-independent date (i.e. July, 1986). This is completely non-informative
   with regards to the patients. So we can suspect a non-informative censoring
   and start investigations using Cox model.

2. What we have to do with the transplant? I.e., `event` has three levels:
   censored, transplant, dead; how we have to consider transplanted patients?
   In this case, clinicians answered that the transplant status is completely
   random! So, we can believe that it is a non-informative censoring.^[In reality
   this is not really true because who stay better is on the top of the list!]
   
Moreover we have to consider, before to start, that `sex` is a categorical
variable, so we have to check (only) the proportionality of the hazards.

```{r}
# Using _survival_ (Cox model for proportional hazard against sex)
cox_sex <- coxph(Surv(time, status == 2) ~ sex,
  data = pbc_df
)

cox.zph(cox_sex)                           # test for proportionality of hazards

cox.zph(cox_sex) %>% 
  plot(
    main = 'Graph of the scaled Schoenfeld residuals for sex, along with a smooth curve',
    col = 'blue'
)
```

The proportional hazard assumption is not invalidate so we can continue with the
analyses.

```{r}
summary(cox_sex)
```

The effect of sex, viewed as hazard ration, say that if you are a female it
seems that you have a lower risk to die, but it is not significant (i.e.,
$p$-value $> 0.05$ and CI include $1$).

> Anyone have the same risk, 1, to die... What the hazard ration say is that if
> at the begin of a day you are alive, if you a re a woman you have 32% less
> probability to die before the end of the day respect a men.

```{r}
# Using rms
rms_sex <- cph(Surv(time, status == 2) ~ sex,
  data = pbc_df,
  x    = TRUE,                    # to compute cox.zph, we need to store x and y
  y    = TRUE
)

cox.zph(rms_sex)

cox.zph(rms_sex) %>% 
  plot(col = 'green')                       # exactly the same results as before!

summary(rms_sex)       # a cleaner and more informative output, note Low and High
```



### Impact of age on death{#age2}

  - we have to check for the proportional HR
  - It is continuous variable, we have to check the the log-linearity too

```{r}
# Using survival
cox_age <- coxph(Surv(time, status == 2) ~ age,
  data = pbc_df
)

cox.zph(cox_age)

cox.zph(cox_age) %>%
  plot(col = 'blue')
```

The proportional hazard hypothesis is not invalidated

> The outputs of rcspline.plot are a plot and a very long matrix with the values
  of xe, lower, upper. The latters are not of our interest, but there are no
  options to not get them. So we include the command into
  `invisible(capture.output(.))` box.^[As suggested by *couthcommander* in
  <https://github.com/CorradoLanera/SuDACDa/issues/2>.]

```{r}
invisible(capture.output(rcspline.plot(
  x       = pbc_df$age,
  y       = pbc_df$time,
  event   = pbc_df$status == 2,
  nk      = 3,                   # default are 5 knots, too mutch for this model
#  model   = 'cox',          # If event is present, model is assumed to be "cox"
  xlab    = 'Age',
  statloc = 'll'
)))
```

The log-linearity is not invalidated

> Note: sometimes you *know* the answer for log-linearity (for any reason), in
>       those cases do not test for it!! (It is not very powerful so for small
>       sample sizes it never reject it)

```{r}
summary(cox_age)
```

The effect is significant but too low to understand, so we can change the
"measure of time" to expand it.


```{r}
coxph(Surv(time, status == 2) ~ I(age / 10),          # consider 10 years as one
  data = pbc_df
) %>%
  summary
```

Here, the effect is increased, but we have to pay attention the an increment of
"one", here, corresponds to an increment of ten years!

```{r}
# Using rms
rms_age <- cph(Surv(time, status == 2) ~ age,
  data = pbc_df,
  x    = TRUE,                    # to compute cox.zph, we need to store x and y
  y    = TRUE
)

Predict(rms_age) %>%
  plot

summary(rms_age)  # _rms_ show effects from the Lower to the Higher limit of IQR
                  # and report the different between them as well as the HR, so
                  # we do not need to perform triky transformation which asks
                  # for an alterate interpretation of the result
```

In particular, the effect quite doubled in fifteen years.^[Good example in which
only the clinicians know if it is an effect clinically relevant (deciding it
__before__ the analyses) or not].


### Impact of aspartate aminotransferase (`ast`) on death {#ast2}
  - same of age
  
```{r}
rms_ast <- cph(Surv(time, status == 2) ~ ast,
  data = pbc_df,
  x    = TRUE,
  y    = TRUE
)

cox.zph(rms_ast)
cox.zph(rms_ast) %>%
  plot(col = 'green')
```

The proportional hazard assumption is not violated, but by the graph it seams
not that linear. Try to transform it using the $log()$ transformation.

```{r}
log_ast <- cph(Surv(time, status == 2) ~ log(ast),
  data = pbc_df,
  x    = TRUE,
  y    = TRUE
)

cox.zph(log_ast)

cox.zph(log_ast) %>%
  plot(col = 'green')
```

The situation is not much better...but we can say that there exists a line
living in the middle of the band...so we are not very happy but we accept it. 

Let's test for log-linearity

```{r}
invisible(capture.output(rcspline.plot(
  x       = log(pbc_df$ast),
  y       = pbc_df$time,
  event   = pbc_df$status == 2,
  xlab    = 'ast',
  statloc = 'll'
)))
```

The log-linear assumption is not violated.

Finally, look at the effect of the log of `ast`

```{r}
summary(log_ast)
```

It is significantly protective, with doubling the effect between the borders of
the IQR.


### Impact of __platelet__ on death {#platelet2}
  - same of age

```{r}
rms_platelet <- cph(Surv(time, status == 2) ~ platelet,
  data = pbc_df,
  x    = TRUE,
  y    = TRUE
)

cox.zph(rms_platelet)

cox.zph(rms_platelet) %>%
  plot(col = 'green')


invisible(capture.output(rcspline.plot(
  x       = pbc_df$platelet %>% as.numeric,
  y       = pbc_df$time,
  event   = pbc_df$status == 2,
  xlab    = 'Platelet',
  statloc = 'll'
)))
```

The log-linear plot has a U-shape so, standard transformation are not good.
We can try to perform a categorization. Two strategy:
  1. look at the log-linear plot and try to find a good cut-points, but we have
    to explain how we have defined them (and "use the p-value" is not a good
    strategy)
  2. Use standard non related cutoff, such as median or quartiles


```{r}
cox_cut_platelet <- coxph(
  Surv(time, status == 2) ~ cut(platelet, c(0, 150, 400, 1000)),
  data = pbc_df
)

cox.zph(cox_cut_platelet)

cox.zph(cox_cut_platelet) %>%
  plot(col = 'blue')

summary(cox_cut_platelet)
```

But here the reference level, i.e. the contrast, is the lower level but the
interested is what happen if we lie above or over the standard values, so we
have to relevel the category to make the medium level as the reference one, i.e.
the first.

```{r}
pbc_df <- pbc_df %>% 
  mutate(
    platelet_ref = cut(pbc_df$platelet, c(0, 150, 400, 1000)) %>%
                      relevel(ref = "(150,400]")
)

cox_relev_platelet <- coxph(Surv(time, status == 2) ~ platelet_ref,
  data = pbc_df
)

summary(cox_relev_platelet)
```



## Investigation on adjusted variables and interactions {#adjusted2}

Clinician: what is the effect of treatment (`trt`) on death?
```{r}
cph(Surv(time, status == 2) ~ trt,
  data = pbc_df
) %>% 
  summary
```

No significant effect for treatment.

Clinician: an adjusted with edema?

```{r}
cph(Surv(time, status == 2) ~ trt + edema,
  data = pbc_df
) %>% 
  summary
```

No effect for treatment nor edema

Clinicians: and what about their interaction?^[The answer here should be "if
there are no marginal significant effect is has no sense to look at the
interaction terms!".]

```{r}
cph(Surv(time, status == 2) ~ trt * edema,
  data = pbc_df
) %>% 
  summary
```

No effect.

Clinicians: and what about adjusted with stage?

```{r}
adj_pbc <- pbc_df %>% 
  mutate(stage_fct = factor(stage))

dd <- datadist(adj_pbc)

cph(Surv(time, status == 2) ~ trt + stage_fct,
  data = adj_pbc
) %>% 
  summary
```

Treatment still with no significant effect. `stage` has some effects, i.e. from
the 3 to 1 or to 4, 

Clinicians: oh, so let's look at the interactions!

```{r}
rms_trt_stage <- cph(Surv(time, status == 2) ~ trt * stage,
  data = adj_pbc
)

summary(rms_trt_stage)

Predict(rms_trt_stage) %>%
  ggplot(anova = anova(rms_trt_stage), pval = TRUE)

Predict(rms_trt_stage) %>%
  ggplot(anova = anova(rms_trt_stage), pval = TRUE)
```

Treatment continue to have no effect

## Longitudinal suvival data analayses {#longitudinal}

Load a data-set, update the `datadist()` for the __rms__ package, and take a 
look at the data

```{r}
pbcseq_df <- as_tibble(pbcseq)
dd <- datadist(pbcseq_df)

pbcseq_df
```

- The only tricky task is to correctly manage and prepare the data. Our proposal
  take advantage of the `dplyr` functionality

```{r}
pbcseq_full <- pbcseq_df %>%
  group_by(id) %>%                    # perform all the next according to the id
  mutate(
    start  = day,                                 # just to have consistent names
    end    = lead(day),           # the end is "the next start" (last will be NA)
    status = if_else(is.na(end), status, 0L), 
    end    = if_else(is.na(end), futime, end) # fill the NA-ends (i.e. the lasts)
                                              # with the real end
)
```


### Impact of bilurubine of death {#bilurubine2}

```{r}
bil_mod <- cph(
  Surv(time = start, time2 = end, event = status == 2L) ~ log(bili),
  data = pbcseq_full,
  x    = TRUE,
  y    = TRUE
)

summary(bil_mod)

plot(
  x   = pbcseq_full$end,
  y   = residuals(bil_mod, type = 'dfbeta'),
  col = 'red'
)
```

### Impact of ast {#ast_full2}

```{r}
ast_mod <- cph(
  Surv(time = start, time2 = end, event = status == 2L) ~ ast,
  data = pbcseq_full,
  x    = TRUE,
  y    = TRUE
)

summary(ast_mod)

plot(
  x   = pbcseq_full$end,
  y   = residuals(ast_mod, type = 'dfbeta'),
  col = 'red'
)

plot(
  x   = pbcseq_full$end,
  y   = residuals(ast_mod, type = 'martingale'),
  col = 'brown'
)
```

What happen with the strange observations? We try to find which is that outlier.

```{r}
# look at the residual caracteristics
residuals(ast_mod, type = 'martingale') %>%
  describe

# take the id of the lowest
strange_id <- residuals(ast_mod, type = 'martingale') %>%
  which.min

# take a look to the ast
pbcseq_full$ast %>% describe

# check the id
pbcseq_full$ast[[strange_id]]
```

Here is another example in which the opinion of a clinician is mandatory, i.e.
we cannot decide if ignore outliers, which ones, etc




## Prognostic model {#prognostic2}

### prognostic model with `àscites`, `edema`, `sex`, `bili`, `ast`, `platelet`, `stage` {#model2}

```{r}
# prepare an ad hoc data frame
pbc_updated <- pbc_df %>% 
  mutate(
    bili_log     = log(bili),
    ast_log      = log(ast),
    platelet_ref = platelet_ref, # we have already defined it
    stage_fct    = as.factor(stage)
  )

dd <- datadist(pbc_updated)

# take a look at them 
pbc_updated %>% 
  dplyr::select(
    ascites, edema, sex, bili_log, ast_log, platelet_ref, stage_fct
  ) %>% 
  describe
```

There are 11 basic df (one each continuous variable and one-minus-n-level for
the categorical one), so to use all of them we need at least $110$ obs. Data has
$418$, this allow us to use a more complex model, with some interaction, splines,
etc (more or less other $15 -- 30$ df).

We decide (following suggestions from @harrell2015regression) to consider
splines for any continuous variable (with 3 knots) and consider sex interaction 
with them and the other numerical variables, leading to near $20$ df.

```{r}
data_used <- pbc_updated %>% 
  dplyr::select(status, time,
    sex, ascites, edema, bili_log, ast_log, platelet_ref, stage_fct
)

dd <- datadist(data_used)


# all the data-set
cph(
  Surv(time, status == 2) ~
    sex * (ascites + edema + rcs(bili_log, 3) + rcs(ast_log, 3)) +
    platelet_ref + stage_fct,
  data = data_used
) %>% 
  summary


# without missing data, and with beckward stepwise variable selection
cph(
  Surv(time, status == 2) ~
    sex + ascites + edema + bili_log + ast_log + platelet_ref + stage_fct,
  data = pbc_updated %>% 
    filter(complete.cases(.))
) %>% 
  step(trace = 0) %>% 
  summary
```

