# *Thusday*: Cox models {#thusday}

```{r data_load}
data(pbc)
?pbc
pbc_df <- as.tibble(pbc)
dd <- datadist(pbc_df)
```


## TP1

1. Impact of __sex__ on death
  - First question: there are non informative censoring. Because there is
    a final date (i.e. July, 1986) this is completerly non informative about
    the patients. So we can suspect a non informative censoring. So we can
    use Cox, KM, ...
  - What we have to do with the transplant? Considercensored, cases, what?
    Answere: it is completely random! So, we can beleve that this is a non 
    informative censoring.^[In reality this is not really true because who
    stay better is on the top of the list!]
  - sex is categorical so we have to check (only) the proportional HR

```{r}
pbc_df

# Using _survival_
cox_sex <- coxph(Surv(time, status == 2) ~ sex,
  data = pbc_df
)

cox.zph(cox_sex)                                      # test for proportional HR

summary(cox_sex)


# Using rms
rms_mod <- cph(Surv(time, status == 2) ~ sex,
  data = pbc_df
)

cox.zph(cox_mod)                                      # test for proportional HR

summary(cox_mod)


```

So, if you are a female it seams that you have a lower risk to die, but it is
not significant (CI include $1$).

2. Impact of __age__ on death

  - we have to check for the proportional HR
  - It is continuous, so we have to check the linearity of the log-linear

```{r}
cox_age <- coxph(Surv(time, status == 2) ~ age,
  data = pbc_df
)

cox.zph(cox_age)                                      # test for proportional HR

cox.zph(cox_age) %>%
  plot                                              # the plot should seams flat

rcspline.plot(
  x     = pbc_df$age,
  y     = pbc_df$time,
  event = pbc_df$status == 2,
  model = 'cox',
  xlab  = 'Age'
)
```

> Note: sometimes you __know__ the answer for log-linearity (for any reason), in
>       those cases do not test for it!! (It is not very powerful so for small
>       ss it never reject it)

```{r}
summary(cox_age)
```

The effect seams quite low, even if significant, so to explain better to a 
clinicians this fact we can find the effect for ten year and not one

```{r}
coxph(Surv(time, status == 2) ~ I(age / 10),
  data = pbc_df
) %>%
  summary
```


3. Impact of __ast__ on death
  - same of age
  
```{r}
cox_ast <- coxph(Surv(time, status == 2) ~ ast,
  data = pbc_df
)

cox.zph(cox_ast)                                      # test for proportional HR

cox.zph(cox_ast) %>%
  plot                                              # the plot should seams flat
```

The proportional HR assumption is violated, so we cannot use a Cox model as it
is, we have to transform it.

```{r}
cox_log_ast <- coxph(Surv(time, status == 2) ~ log(ast),
  data = pbc_df
)

cox.zph(cox_log_ast)

cox.zph(cox_log_ast) %>%
  plot                                              # the plot should seams flat
```
There exixt a line live in the middle of the band...so we are not very happy
but we accept it. 

Let's test for log-linearity

```{r}
rcspline.plot(
  x     = log(pbc_df$ast),
  y     = pbc_df$time,
  event = pbc_df$status == 2,
  model = 'cox',
  xlab  = 'AST'
)
```

And finally look at the effect of the log of `ast`
```{r}
summary(cox_ast)
```

3. Impact of __platelet__ on death
  - same of age

```{r}
cox_platelet <- coxph(Surv(time, status == 2) ~ platelet,
  data = pbc_df
)

cox.zph(cox_platelet)

cox.zph(cox_platelet) %>%
  plot                                              # the plot should seams flat

rcspline.plot(
  x     = pbc_df$platelet %>% as.numeric,
  y     = pbc_df$time,
  event = pbc_df$status == 2,
  model = 'cox',
  xlab  = 'Platelet'
)
```

The log-linear plot has a U-shape so, standard transformation are not good.
We can try to perform a categorization. Two startegy:
  - look at the log-linear plot and try to find a good cutpoints, but we have
    to explain how we have defined them (and "use the p-value" is not a good
    strategy)
  - Use standard non related cutof, such as median or quartile


```{r}
cox_cut_platelet <- coxph(Surv(time, status == 2) ~ cut(platelet, c(0, 150, 400, 1000)),
  data = pbc_df
)

cox.zph(cox_cut_platelet)

cox.zph(cox_cut_platelet) %>%
  plot                                              # the plot should seams flat

summary(cox_cut_platelet)
```

But here the reference is the lower level but the interested is what happen if
we lie above or over the standard values, so we have to relevel the category
to make the medium level as the reference one, i.e. the first

```{r}
pbc_df <- pbc_df %>% 
  mutate(
    platelet_ref = cut(pbc_df$platelet, c(0, 150, 400, 1000)) %>%
                      relevel(ref = "(150,400]")
)

cox_relev_platelet <- coxph(Surv(time, status == 2) ~ platelet_ref,
  data = pbc_df
)

summary(cox_relev_platelet)
```



## TP2

```{r}
coxph(Surv(time, status == 2) ~ trt,
  data = pbc_df
) %>% 
  summary

coxph(Surv(time, status == 2) ~ trt + edema,
  data = pbc_df
) %>% 
  summary

coxph(Surv(time, status == 2) ~ trt * edema,
  data = pbc_df
) %>% 
  summary

coxph(Surv(time, status == 2) ~ trt + stage,
  data = pbc_df %>% mutate(stage = factor(stage, ordered = TRUE))
) %>% 
  summary

coxph(Surv(time, status == 2) ~ trt * stage,
  data = pbc_df %>% mutate(stage = as.factor(stage))
) %>% 
  summary


Predict(cph(Surv(time, status == 2) ~ trt * stage,
  data = pbc_df %>% mutate(stage = as.factor(stage))
)) %>%
  ggplot
```



## TP3

```{r}
pbcseq_df <- as.tibble(pbcseq)
dd <- datadist(pbcseq_df)

pbcseq_df
```


- impact of bilurubine of death

```{r}
pbcseq_full <- pbcseq_df %>%
  group_by(id) %>%                    # perform all the next according to the id
  mutate(
    start  = day,                                 # just to have consistent names
    end    = lead(day),           # the end is "the next start" (last will be NA)
    status = if_else(is.na(end), status, 0L), 
    end    = if_else(is.na(end), futime, end) # fill the NA-ends (i.e. the lasts)
                                              # with the real end
)
```
```{r}
bil_mod <- coxph(Surv(time = start, time2 = end, event = status == 2L) ~ log(bili),
  data = pbcseq_full
)

summary(bil_mod)

plot(
  x = pbcseq_full$end,
  y = residuals(bil_mod, type = 'dfbeta')
)
```



- impact of ast
```{r}
ast_mod <- coxph(Surv(time = start, time2 = end, event = status == 2L) ~ ast,
  data = pbcseq_full
)

summary(ast_mod)

plot(
  x = pbcseq_full$end,
  y = residuals(ast_mod, type = 'dfbeta')
)

plot(
  x = pbcseq_full$end,
  y = residuals(ast_mod, type = 'martingale')
)
```

What happen with the strange observations? I  try to find which is the outlier

```{r}
residuals(ast_mod, type = 'martingale') %>%
  describe

strange_id <- residuals(ast_mod, type = 'martingale') %>%
  which.min

pbcseq_full$ast %>% describe

pbcseq_full$ast[[strange_id]]
```



## TP4

1. priognostic model with
  - `àscites`, `edema`, `sex`, `bili`, `ast`, `platelet`, `stage`

```{r}
pbc_updated <- pbc_df %>% 
  mutate(
    bili_log     = log(bili),
    ast_log      = log(ast),
    platelet_ref = platelet_ref, # we have already defined it
    stage_fct    = as.factor(stage)
  )

dd <- datadist(pbc_updated)

pbc_updated %>% 
  dplyr::select(
    ascites, edema, sex, bili_log, ast_log, platelet_ref, stage_fct
  ) %>% 
  describe
# there are 11 df so to use all of them we need at least 110 obs


coxph(
  Surv(time, status == 2) ~
    sex * (ascites + edema + rcs(bili_log, 3) + rcs(ast_log, 3)) + platelet_ref + stage_fct,
  data = pbc_updated
) %>% 
  summary


coxph(
  Surv(time, status == 2) ~
    sex + ascites + edema + bili_log + ast_log + platelet_ref + stage_fct,
  data = pbc_updated %>% 
    dplyr::select(status, time,
      sex, ascites, edema, bili_log, ast_log, platelet_ref, stage_fct
    ) %>% 
    filter(complete.cases(.))
) %>% 
  step(trace = 0)
```

