# *Wednesday*: Competing risk {#competing-risk}

## Key (operative) concepts {#key3}

1. Patient are exposed simultaneously to $k (\geq2)$ causes

2. Effect Free Survival (EFS) is univariate, i.e. only the First Observed Event
   (FOE) is considered and of interest

3. The interest is not in the survival model

> "At $\infty$ all  individuals will not die in the ICU"

4. Type of observed time
  - Cansored (conventionally coded w/ $0$)
  - Failure w/ a FOE different from the last absorbing one (coded w/ $1 -- k-1$)
  - Failure w/ the FOE as the last absorbing event (coded w/ $k$)
  - $T_k = min\Bigl(\tilde{T_k}^{D_k} |
      D_k \in \{\textrm{causes of failure for }k\}
    \Bigr)$

5. Cumulative Incidence Function (CIF) do not require independence between
   causes

6. In competing risk, K-M is biased, i.e. overestimates the CIF (because it the
   independence assumption is violated)

7. Tests
  - w/out competing risk: log-rank
  - w/ competing risk: modified $\chi^2$ [@gray1988class]

8. Regression strategies for competing risk
  - Case Specific Hazard Ratio (CS-HR) --- Cox, useful for clinical interests (present it for each competing risk taken singularly)
  - Subdistribution Hazard Ratio (SHR) --- Fine-Gray, useful for administrative]
    interests (present it for the global risk considering the contribution of
    each competing one)

> Test the proportional hazard assumption for SHR

> There are formulas for the sample size calculation when considering competing
  risk

## Data manipulation

```{r}
set.seed(171004)
data(mgus, package = 'survival')
# ?mgus

mgus_df <- as_tibble(mgus)
dd <- datadist(mgus_df)

mgus_df
```


1. Find number of patient w/ malignancy (AKA transition), death
   (w/out malignancy) and Free of Events.

```{r}
mgus_df <- mgus_df %>% 
  mutate(
    malignancy = !is.na(pcdx)
  )

mgus_df %>% 
  group_by(malignancy, death) %>% 
  summarise(n = n())
```

Patients w/ malignancy as a FOE are 64; patients which experiment death as FOE are 163, while the ones FoE are 14.
163.

2. Find the indicator for censored, malignancy and death (`indicator`)
3. Find the time-to-event to use in the models (`time_t`)

```{r}
mgus_df <- mgus_df %>% 
  mutate(
    indicator = if_else(malignancy, 1, 2 * death),
    time_t    = pmin(futime, pctime, na.rm = TRUE)
)

mgus_df
```

4. Estimate the naive K-M and the cumulative incidence functions

```{r}
# Using survival
cuminc(mgus_df$time_t, mgus_df$indicator) %>%                  # ?cmprsk::cuminc
  plot(                                                  # ?cmprsk:::plot.cuminc
    main    = 'Cumulative Incidence Estimates curves',
    col     = c('blue', 'red'),
    xlab    = 'Days',
    curvlab = c('Transition', 'Death'),
    wh      = c(1, 1)                                          # legend position
)

survfit(Surv(time_t, malignancy) ~ 1,        # using `rms::npsurv()` is the same
  data = mgus_df
  ) %>% 
  lines(                          # Use `lines()` to draw over the previous plot
    fun       = 'event',                            # plot the cumulative events
    conf.int   = FALSE,
    col       = 'black',
    lty       = 3
)
legend(x = 1, y = 0.86,
  legend = 'Naive K-M (Overestimation of Transitions)',
  col    = 'black',
  lty    = 3,
  bty    = 'n' # remove box arround the legend (because we have to add an entry)
)
```



## Simulation of Competing risk

1. Specify two cause-specific exponential hazard $\lambda_1(t)$ and
$\lambda_2(t)$ of means $0.8$ and $1.2$. (Set sample size as you like.)

```{r}
n        <- 1e4
lambda_1 <- 0.8
lambda_2 <- 1.2
```


2. Simulate survival times $T$ based on the all causes hazard
   $\lambda_.(t) = \lambda_1(t) + \lambda_2(t)$.

```{r}
lambda    <- lambda_1 + lambda_2 
surv_time <- rexp(n,
  rate = 1 / lambda
)
```


3. Generate Bernoulli $B(p)$ random variables, w/
   $p = \lambda_1(t) / \lambda_.(t)$, i.e. is the probability of occurrence of
   the event of type 1.

```{r}
p_cens     <- lambda_1 / lambda
transition <- rbinom(n,
  size = 1,
  prob = p_cens
) %>%
  as.logical     # Set as logical to use the variable for conditional statements
```


4. Simulate uniform censoring times over $[0, 1]$.

```{r}
censor_time <- runif(n,
  min = 0,
  max = 1
)
```


5. Estimate the Cumulative Incidence of each competing event, w/ and w/out
   censoring; discuss the results.

```{r}
# create the dataset
sim_data <- data_frame(
  id         = seq_len(n),
  transition = transition,
  surv_t     = surv_time,
  cens_t     = censor_time,
  time_t     = pmin(surv_t, cens_t),
  status     = case_when(
    time_t == cens_t ~ 0L,              # All the censored patients has status 0
    transition       ~ 1L,    # Among the other, the ones which has a transition
                              # have state 1
    TRUE             ~ 2L     # All the other were dead (before the end of f-up)
  )
)

# Explore a (random) sample of three cases for each staus
sim_data %>% 
  group_by(status) %>% 
  sample_n(3)
```

```{r}
# Using survival
cuminc(sim_data$time_t, sim_data$status) %>%                  # ?cmprsk::cuminc
  plot(                                                  # ?cmprsk:::plot.cuminc
    main    = 'Cumulative Incidence Estimates curves',
    col     = c('blue', 'red'),
    xlab    = 'Time (normalized [0, 1])',
    curvlab = c('Transition', 'Event'),
    wh      = c(0.01, 1)                                          # legend position
)

survfit(Surv(time_t, transition) ~ 1,        # using `rms::npsurv()` is the same
  data = sim_data
  ) %>% 
  lines(                          # Use `lines()` to draw over the previous plot
    fun       = 'event',                            # plot the cumulative events
    conf.int   = FALSE,
    col       = 'black',
    lty       = 3
)
legend(x = 0.01, y = 0.86,
  legend = 'Naive K-M (overestimation of transitions)',
  col    = 'black',
  lty    = 3,
  bty    = 'n' # remove box arround the legend (because we have to add an entry)
)

```



## Estimation of the effect of sex on MGUS incidence

1. Compare the results of Cox cause specific hazard model...

> For clinical questions, i.e. cause specific risk to experiment the event
  w/out taking into account the other couse(s)

```{r}
dd <- datadist(mgus_df)

cox_sex <- cph(Surv(time_t, malignancy) ~ sex,
  data = mgus_df,
  x    = TRUE,
  y    = TRUE
)

summary(cox_sex)                  # this is good for a clean view of the effects
cox_sex                    # Here there are more informations (and the p-values)

Predict(cox_sex) %>%      # It is necessary to have the predictions for the plot
  plot


cox_sex_death <- cph(Surv(time_t, indicator == 2) ~ sex,
  data = mgus_df,
  x    = TRUE,
  y    = TRUE
)

summary(cox_sex_death)
cox_sex_death

Predict(cox_sex_death) %>% 
  plot
```


2. ...to those of the Fine and Gray model

> For administrative questions, i.e. overall risk of experiment each event
  taking into account the competing risk

```{r}
mgus_num <- mgus_df %>% 
  mutate(sex = as.numeric(sex))

crr(                                      # We do not have a plot method for crr
  ftime   = mgus_num$time_t,
  fstatus = mgus_num$indicator,
  cov1    = mgus_num$sex
) %>%
  summary
```

