--- 
title: "Survival Data Analysis for Cancer Data"
author: "Corrado Lanera, Danila Azzolina, Daniele Bottigliengo^[[Unit of Biostatistics, Epidemiology and Public Health](http://www.dctv.unipd.it/dipartimento/strutture/biostatistica) of the [Dep. of Cardiac, Thoracic and Vascular Sciences](http://www.dctv.unipd.it/) --- [Univ. of Padova](http://www.unipd.it/)]"
date: "02 - 06 October, 2017"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
url: https\://corradolanera.github.io/SuDACDa/
github-repo: CorradoLanera/SuDACDa
description: "Examples and exercise from the Politecnico Ph.D. course by prof. Sylvie Chevret and prof. Matthieu Resche-Rigon."
---

# Introduction {-}

This book is designed to collect notes and exercises from the Ph.D. course
on [__Survival Data Analysis for Cancer Data__](http://www.politocomunica.polito.it/events/appuntamenti/(idnews)/9665)
by prof. Sylvie Chevret
and prof. Matthieu Resche-Rigon from
[ECSTRRA Team](http://www.cress-umr1153.fr/),
[Inserm](https://www.inserm.fr/),
[University of Paris Diderot](https://www.univ-paris-diderot.fr/), promoted
by the
[Dep. of Mathematical Sciences "G. L. Lagrange"](http://www.disma.polito.it/) of
the [Politecnico](http://www.polito.it/) of Torino (Italy).


## Contributions {-}

Any contribution is welcome! From the download button on the top of each (HTML) page you can download both the `epub` and the `PDF` versions of the present book. 


If you find any mistake/typo or want to share ideas, you can help improve the book in the following way:

  - Providing a solution proposal by opening a pull request to the related git repository  (<https://github.com/CorradoLanera/SuDACDa/pulls>)
  
  - Asking for a fix by opening an issue to the project
   (<https://github.com/CorradoLanera/SuDACDa/issues>)




## Settings {-}

Here, there are the libraries loaded during the course, w/ the relative options,
plus some packages and options useful to write code more understandable by
humans obtaining nicer output.

```{r setup, message=FALSE, warning=FALSE}
# Packages for the analyses
library(survival)                                            # Survival Analysis
library(survminer)                     # Drawing Survival Curves using 'ggplot2'
library(cmprsk)                                                 # Competing risk
library(rms)              # Regression Modeling Strategy (include Hmisc package)
  options(datadist = 'dd')                  # Distribution Summaries used by rms

# Package(s) for data management 
library(tidyverse)                    # Imports the principal tidyverse packages

# Document output options
knitr::opts_chunk$set(
	echo        = TRUE,                                      # Render all the code
	message     = FALSE,                                  # Do net render messages
	warning     = FALSE,                                  # Do not render warnings
	fig.height  = 4.4,   # Right figure height to permit two figures in a PDF page
	cache.extra = knitr::rand_seed   # cache random seed to assure reproducibility
)
```

The following code create the packages.bib files which is the BibTeX lists of
all the packages references we have loaded.

```{r}
# Automatically create a bib database for the loaded packages
knitr::write_bib(c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
  file = 'packages.bib'
)
```
