--- 
title: "Survival Data Analysis for Cancer Data"
author: "Corrado Lanera, Danila Azzolina and Daniele Bittigliengo^[[Unit of Biostatistics, Epidemiology and Public Health](http://www.dctv.unipd.it/dipartimento/strutture/biostatistica) of the [Dep. of Cardiac, Thoracic and Vascular Sciences](http://www.dctv.unipd.it/) --- [Univ. of Padova](http://www.unipd.it/)]"
date: "02 - 06 October, 2017"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: CorradoLanera/SuDACDa
description: "Examples and exercise from the Politecnico of Torino (Italy) Ph.D. course on Survival Data Analysis for Cancer Data."
---

# Introduction {-}

In this book, there are our notes and exercices from the Ph.D. course
on __Survival Data Analysis for Cancer Data__ by Prof. Matthieu Resche-Rigon and
Prof. Sylvie Chevret from [ECSTRA Team](http://www.cress-umr1153.fr/),
[Inserm](https://www.inserm.fr/),
[University of Paris Diderot](https://www.univ-paris-diderot.fr/), promoted
by the
[Dep. of Mathematical Sciences "G. L. Lagrange"](http://www.disma.polito.it/) of
the [Politecnico](http://www.polito.it/) of Torino (Italy).





## Contributions {-}

If you find any mistakes, typing errors or if you simply want to contribute, you have two main options:

1. Provide a solution proposal by opening a *pull request* to the related git
   repository (<https://github.com/CorradoLanera/SuDACDa/pulls>)
2. Ask us fo a fix by opening an *issue* to the project
   (<https://github.com/CorradoLanera/SuDACDa/issues>)

From the download button on the top of each (HTML) page you can download both
the `epub` and the `PDF` versions of the present book. 




## Settings {-}

Here, there are the libraries loaded during the course, with the relative
options, plus some packages and options useful to write code more understandable
by humans obtaining nicer output.

```{r setup, message=FALSE, warning=FALSE}
# Packages for the analyses
library(survival)                                            # Survival Analysis
library(survminer)                     # Drawing Survival Curves using 'ggplot2'
library(rms)                                      # Regression Modeling Strategy
  options(datadist = 'dd')                  # Distribution Summaries used by rms

# Packages for data management 
library(tidyverse)                    # Imports the principal tidyverse packages

# Document output options
knitr::opts_chunk$set(
	echo       = TRUE,
	message    = FALSE,
	warning    = FALSE,
	fig.height = 4.4
)
         # by default, render all the code too
```

The following code create the packages.bib files which is the bibtex lists of
all the packages references we have loaded.

```{r}
# Automatically create a bib database for the loaded packages
knitr::write_bib(c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
  file = 'packages.bib'
)
```
